---
title: "STdata package overview"
author: 
  - name: Lukas M. Weber
    affiliation: "Johns Hopkins Bloomberg School of Public Health"
package: STdata
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{STdata package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

The `STdata` package provides access to a collection of spatially resolved transcriptomics datasets, which have been formatted into the [SpatialExperiment](https://bioconductor.org/packages/SpatialExperiment) Bioconductor object class.

These datasets have been collected from various publicly available sources, and cover several technological platforms. We provide them in the form of `SpatialExperiments` objects to make them easier to access, so that we and others can use them for examples, demonstrations, tutorials, and other purposes.

The `SpatialExperiment` class is an extension of `SingleCellExperiment`, adapted for the properties of spatially resolved transcriptomics data.


# Datasets

The package contains the following datasets:

- `mouse_coronal` (10x Genomics Visium): a single coronal section from the mouse brain, spanning one hemisphere. Previously provided by 10x Genomics from their [website](https://support.10xgenomics.com/spatial-gene-expression/datasets/1.1.0/V1_Adult_Mouse_Brain).

- `human_DLPFC` (10x Genomics Visium): a single sample (sample 151673) from our previously published dataset consisting of 12 samples from the human dorsolateral prefrontal cortex (DLPFC) in the human brain, from 3 neurotypical donors. The original dataset is available via the [spatialLIBD](http://bioconductor.org/packages/spatialLIBD) Bioconductor package, and is described in our paper [Maynard and Collado-Torres et al. 2020](https://www.biorxiv.org/content/10.1101/2020.02.28.969931v1).


# Load data

The following example shows how to load the `SpatialExperiment` data objects in an R session.

```{r, message=FALSE}
library(STdata)
library(SpatialExperiment)

# load SpatialExperiment object
spe <- load_data("mouse_coronal")
spe

# inspect object
assayNames(spe)
rowData(spe)
colData(spe)
spatialCoords(spe)
imgData(spe)
```

